---
title: "Trump vs. Biden 2020: Logistic Regression and Post-Stratification Analysis"
author: "Natasha Gulati"
date: "November 5, 2020"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

```{r setup, include=FALSE}
library(tidyverse)

# Loading in the cleaned survey Data
survey_data <- read_csv("C:/Users/User/Desktop/PS3/outputs/survey_data.csv")

# Loading in the cleaned census Data
census_data <- read_csv("C:/Users/User/Desktop/PS3/outputs/census_data.csv")

```

# Abstract
Using post-stratification on a logistic regression model I derive the probability of Joe Biden winning the 2020 Presidential Election based on how each state votes. Electoral votes dictate a lot of importance when it comes to winning the presidential race and calculating the estimate and predictor values will aide in the help of understanding who will win the election. Post-stratifying the data, expressing summary tables and coefficient plots, I analyze the predictions of the 2020 election census and survey data.


# Model
Here we are interested in predicting the popular vote outcome of the 2020 American federal election. To do this we are employing a post-stratification technique. In the following sub-sections I will describe the model specifics and the post-stratification calculation. We will be using the census data from IPMUS (Ruggles, Flood and peers, 2020) and the survey data from 
Democracy Fund + UCLA Nationscape (Tausanovitch, Chris and Lynn Vavreck, 2020). They are based on the feelings of voters from across the United States that may vote in the 2020 Presidential election.

## Model Specifics
I will use the logistic regression model to model the proportion of voters who will vote for Joe Biden because the outcome of the logistic regression model is binary, if the vote is a yes (1) or a no (0). I will be using the votes for Biden based on the state in which they live in to model the probability of voting for Joe Biden. The census data was sorted and grouped by state. The logistic regression model for this paper will be:

$$ log(p/1−p) = \beta_0+\beta_1  x_{AL}+\beta_2  x_{AR}+... +\beta{_j}  x_{state}+  \epsilon $$ where $j = 1,...,51$ and $state = AL,...,WY$ for all 51 states.

The proportion $p$ represents the probability of voters who will vote for Joe Biden. $\beta_0$ represents the intercept of the model where votes answered NA to the study question asking for state of residence. It is a negative value so we can negate the value on a scatter plot, but it does indicate that those who did not indicate their state are less likely to vote for Biden. Additionally, $\beta_j$ coefficients represent change in log odds. So, for everyone one unit increase in vote within a state, we expect the log odds of $\beta_j$ to increase the probability of voting for Joe Biden of the corresponding $x_{state}$. A negative estimate value would be an indicator of the state's likelihood of voting for Donald Trump.

# Summary of the Logistic Model
```{r, echo=FALSE}

# Creating the Model

model <- glm(vote_trump ~ state, 
            data=survey_data, family= "binomial")
model2 <- glm(vote_biden ~ state , 
            data=survey_data, family= "binomial")
summary(model2)
## Survey Estimation for Logistic Regression, Model Results (to Report in Results section)
#install.packages("jtools")
library(jtools)

broom.mixed::tidy(model2)
a = jtools::summ(model2, scale= TRUE)
a

```

## Post-Stratification 

In order to estimate the proportion of voters who will vote for Joe Biden I will perform a post-stratification analysis. Each cell will be grouped from voters from each state. Using the model described in the previous sub-section, I will estimate the proportion of voters in each state's bin, which will be the value of $y_j$. I will then weight each proportion estimate (within each bin) by the respective population size of that bin, which is $N_j$ and sum those values and divide that by the entire population size, $N_j$, to get the proportion of voters who will vote for Joe Biden, $y^{PS}$ The formula is as follows:
$$ y^{PS}= \sum N_jy_j/N_j$$ 
$ y^{PS}= 0.418$ which means that the probability of Joe Biden winning the 2020 Presidential Election based on voters in their respective states is around 42%. Calculating $y^{PS}$ for whether Donald Trump (using the same logistic regression model), $ y^{PS}= 0.392$ This means that the proportion of voters over all to vote for Trump is around 39%. Thus, the Democratic party is predicted to win the 2020 election, though the difference is not statistically strong.

```{r, include=FALSE}

# Here I will perform the post-stratification calculation
census_data$logodds_estimate <-
  model2 %>%
  predict(newdata = census_data)
# Creating y_j
census_data$estimate <- exp(census_data$logodds_estimate)/(1+exp(census_data$logodds_estimate))
# Creating a column that is y_j multiplied by N_j of each cell called jb_predict_prop

census_data$jb_predict_prop = census_data$estimate*census_data$n

# Summation of jb_predict_prop
a = c(census_data$jb_predict_prop)

j = sum(a, na.rm = TRUE)

# Summation of jb_predict_prop divided by summation of N_j to get y^PS

jb_predict = j/sum(census_data$n)

jb_predict


#For $y^PS for Donald Trump votes


census_data$logodds_estimate <-
  model %>%
  predict(newdata = census_data)
# Creating y_j
census_data$estimate <- exp(census_data$logodds_estimate)/(1+exp(census_data$logodds_estimate))
# Creating a column that is y_j multiplied by N_j of each cell called jb_predict_prop

census_data$d_predict_prop = census_data$estimate*census_data$n

# Summation of d_predict_prop
a = c(census_data$d_predict_prop)

j = sum(a, na.rm = TRUE)

# Summation of d_predict_prop divided by summation of N_j to get y^PS

d_predict = j/sum(census_data$n)

d_predict

```


 

# Results


### Table 1
```{r, echo=FALSE}
smoke <- matrix(c(-5.406, 0.881,0.951,4.519,	0.822,0.582, 8.046,	0.829, 0.332, 9.99, 0.829, 0.228, 3.834,	0.826, 0.642, 9.650,	0.836, 0.248,	8.005,	0.821, 0.329),ncol=3,byrow=TRUE)
colnames(smoke) <- c("Estimate","Standard Error","p-value")
rownames(smoke) <- c("AR","TX","GA", "MI", "PA", "WI", "FL")
smoke <- as.table(smoke)
knitr::kable(smoke)
```
Above we have the estimate, standard error and p-value for Arizona, Texas, Georgia, Michigan, Pennsylvania, Wisconsin and Florida which are commonly known as "Battle Ground" or "swing" states. This means they are not pre-disposed to be either a red or blue state, but rather can 'swing' either way. It is widely known that these states can make or break an election due to the value of their electoral votes (Forsberg and Payton, 2015).


### Table 2
```{r, echo=FALSE}
ttbl <- matrix(c( 0.03,8777.89, 9123.45),ncol=1,byrow=TRUE)
colnames(ttbl) <- c("Values")
rownames(ttbl) <- c("R^2","AIC","BIC")
ttbl <- as.table(ttbl)
knitr::kable(ttbl)

```
The table consists of the R^2, AIC and BIC value of logistic regression model on Biden voters based on their state.


### Figure 1 : Logistic Regression coefficient plot of Joe Biden votes in each state:
```{r, echo=FALSE}
require(jtools)
jtools::plot_coefs(model2, ci_level = 0.95, scale = TRUE)

```
Here the plot is based on a confidence interval of 95%.

### Figure 2:  Logistic Regression coefficient plot of Donald Trump votes in each state:
```{r, echo=FALSE}
require(jtools)
jtools::plot_coefs(model, ci_level = 0.95, scale = TRUE)
```

Here we have the estimate plots for the same logistic model, but for those who would vote for Donald Trump based on the same confidence interval of 95%.


# Discussion

We see that the post-stratification model based on the state in which a voter lives predicts a likelihood of 42% that people will vote for Joe Biden and the Democratic party will win. When we view the estimates in the summary of the model we can see that majority of the estimates are positive values with similar standard errors for each state. If we look at Table 1, we can see a select few states that are regarded as swing states (Forsberg and Payton, 2015) these states are important to discuss as they frame the probability of votes within those states and are regarded as tie-breakers. As we can see, Georgia, Michigan, Florida and Wisconsin have high estimates (~8-9) and low p-values (~0.2-0.3) so we can say with a moderate likelihood that Joe Biden has a high chance of winning in GA, MI, FL and WI based on the data. Still looking at Table 1, Arizona has a negative slope and a very high p-value (~0.95), we can say that Joe Biden is strongly not likely to win Arizona. Lastly, we can see that Pennsylvania has a lower positive slope and a p-value of around 0.6. This does not have a high statistical significance as the strength of the vote for either way, but it weakly projected that Biden will win in PA.

In Table 2, R^2, AIC and BIC values are calculated for the logistic regression model. The very low R^2 value indicates that the independent variable does not explain the outcome of the results even though 'state' variable is significant. This means that the variation in my dependent data is not explained by the 'state' explanatory variable, so the state one resides in therefore does not impact if they will vote for Joe Biden or Donald Trump. AIC and BIC are information criterion measurements that essentially tell you if a model is benefited with the number of parameters used and the type of model used. They are both extremely high, and should be around 2-6 to be significantly optimal. With this, we can say that the use of a single logistic predictor variable is not the best to thoroughly analyze the data, and possibly the use of additional parameters would be beneficial.

In Figure 1, we have a Logistic Regression coefficient plot of Joe Biden votes in each state, and it is evident that majority of each estimate for each state is positive and a vertical trend appears around positive 1. Within Figure 2, the same model is displayed but the binomial probability tallied was votes for Trump. There is almost a reflection of the data in Figure 2 from Figure 1. This is understandable as the binomial factor of $p$ being the probability of someone voting for Biden, $1-p$ is the probability of them voting for Trump.


# Weaknesses

There are a lot of other variables or weights I could have used when initiating this analysis. Touching on the R^2, AIC, and BIC values mentioned above, there are better models and parameters that could be used to explain the results of the report. For example, the prevalence of race, sex and education play a large part of *why* someone would vote for Biden or for Trump. Race is a heavy and vastly important topic in the United States, as well as where one is on the social status ladder or what their income level is. Since income levels are related to taxes, and taxes are a large part of *why* one would possibly vote Democratic or Republican. The input of alternate or additional parameters would possibly give us a lower AIC, BIC and a higher R^2 levels, which would give us a better explanation of why Biden or Trump would have been voted for.



## Next Steps

Through the high AIC and low R^2 values, we can say that the logistic regression model was not the most optimal use to analyze the data. The use of a multilevel Bayesian or Frequentist model could be used in order to account for more variables and in-depth reasoning to the outcome of the data. The use of state being a level 1 group and education, race, sex etc. could be level 2 grouping in a multilevel model. Moving forward to clarify the results in this study, the use of a survey of the same variables/questions asked but instead we tally the actual results of who actually voted for Trump or Biden. Since the data was collected prior to the election, cross-referencing it to post-election data (post-hoc analysis) would be beneficial to see whether or not the outcome of this report is consistent with the outcome of the election by state. More so, a deep analysis of the swing states seems necessary for how the results of the election turn out. Possibly on any other further surveys, a question that refers to why the participant voted for the candidate (i.e. taxes, immigration policies etc.) would also be beneficial to the reports analysis.


# References

1. Tausanovitch, Chris and Lynn Vavreck. 2020. Democracy Fund + UCLA Nationscape, October 10-17, 2019 (version 20200814). Retrieved from https://www.voterstudygroup.org/downloads?key=92a1633f-3054-4265-8378-54cc1aa6b351.

2. Steven Ruggles, Sarah Flood, Ronald Goeken, Josiah Grover, Erin Meyer, Jose Pacas and Matthew Sobek. IPUMS USA: Version 10.0 [dataset]. Minneapolis, MN: IPUMS, 2020. https://doi.org/10.18128/D010.V10.0

3. Ole J. Forsberg & Mark E. Payton. 2015. Analysis of Battleground State Presidential Polling Performances, 2004–2012, Statistics and Public Policy, 2:1, 1-10, DOI: 10.1080/2330443X.2015.1034389
